An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-29 EconGrowth-R-Functions.Rnw
Went through all of Caleb's text. Lots of questions remain.

########################
# Creates a lattice plot that has the size of confidence intervals on the vertical axis and parameter
# values on the horizontal axis.
# At this point, this function is NOT a general-purpose function, as the number of permutations
# of model, parameter, energy type, and factor is too varied to be able to turn into a general-purpose 
# graph with little time investment at this point.
# This function makes a lattice plot with the following panels (in the arrangement shown)
#
# gammaCDeq   gammaCDex    gammaCDeu
# deltaCESeq  deltaCESex   deltaCESeu
# sigmaCESeq  sigmaCESex   sigmaCESeu
#
# Parameters and models (gamma for CDe and delta and sigma for CESe) are in rows and energyTypes (q, x, and u) 
# are in columns.
#
# textScaling: the amount by which you want to scale the text on the graph.
##
CIvsParamLattice <- function(textScaling=1.0, countryAbbrevScaling=1.0){
  # Set up plotting parameters that are common to both graphs.
  xTics <- c(0.0, 0.5, 1.0) #x axis tic locations
  yTics <- c(0.0, 1.0, 2.0) #y axis tic locations
  xLimits <- c(-0.2, 1.2) #x axis limits
  yLimits <- c(-0.4, 2.4) #y axis limits
  xLabels <- c("$\\gamma$", "$\\delta$", "$\\sigma$") #x axis labels
  yLabel <- "Width of 95\\% Confidence Interval"
  stripLabels <- c("CD with $q$", "CES with $q$", "CES with $q$",
                   "CD with $x$", "CES with $x$", "CES with $x$", 
                   "CD with $u$", "CES with $u$", "CES with $u$")
  stripBG <- "transparent"

@@

  # Diagonal line for the sigma-CES-Q panel  
  param <- c(0.0, 1.0)
  CI <- c(0.0, 2.0)
  factor <- c("sigmaCESeQ", "sigmaCESeQ")
  diagLineSigmaCESeQDF <- data.frame(param, CI, factor)
  # Diagonal line for the sigma-CES-X panel  
  factor <- c("sigmaCESeX", "sigmaCESeX")
  diagLineSigmaCESeXDF <- data.frame(param, CI, factor)
  # Diagonal line for the sigma-CES-U panel  
  factor <- c("sigmaCESeU", "sigmaCESeU")
  diagLineSigmaCESeUDF <- data.frame(param, CI, factor)

@@

  # Make an xyplot that contains the diagonal lines.
  diagLinesLattice <- xyplot(CI ~ param | factor, data=diagLinesDF,
                             scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                             tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                             alternating=FALSE, # eliminates left-right, top-bot alternating of axes
                             x=list(at=xTics),
                             y=list(at=yTics)
                             ), 
                             xlim=xLimits, 
                             ylim=yLimits, 
                             xlab=list(label=xLabels),
                             ylab=list(label=yLabel),
                             strip = strip.custom(factor.levels=stripLabels, # Set text for factor levels
                                                  bg=stripBG,
                                                  par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                                  ),
                             as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                             )

  # Build a data.frame that has all the necessary data from the models.
  gammaCDeqDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="Q")
  deltaCESeqDF <- CIvsParamDF(model="CESe", param="delta", energyType="Q")
  sigmaCESeqDF <- CIvsParamDF(model="CESe", param="sigma", energyType="Q")
  gammaCDexDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="X")
  deltaCESexDF <- CIvsParamDF(model="CESe", param="delta", energyType="X")
  sigmaCESexDF <- CIvsParamDF(model="CESe", param="sigma", energyType="X")
  gammaCDeuDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="U")
  deltaCESeuDF <- CIvsParamDF(model="CESe", param="delta", energyType="U")
  sigmaCESeuDF <- CIvsParamDF(model="CESe", param="sigma", energyType="U")
  # Change the name of the first columns (the ones that contain the parameter of interest to us) to "param".
  # Doing this allows us to combine the individual data frames into one large data frame.
  colnames(gammaCDeqDF)[1]  <- "param"
  colnames(deltaCESeqDF)[1] <- "param"
  colnames(sigmaCESeqDF)[1] <- "param"
  colnames(gammaCDexDF)[1]  <- "param"
  colnames(deltaCESexDF)[1] <- "param"
  colnames(sigmaCESexDF)[1] <- "param"
  colnames(gammaCDeuDF)[1]  <- "param"
  colnames(deltaCESeuDF)[1] <- "param"
  colnames(sigmaCESeuDF)[1] <- "param"
  # Create one big data.frame with all of the data.
  data <- rbind(gammaCDeqDF, deltaCESeqDF, sigmaCESeqDF, 
                gammaCDexDF, deltaCESexDF, sigmaCESexDF,
                gammaCDeuDF, deltaCESeuDF, sigmaCESeuDF)
print(data)
  # Now make the graph.
  # Need to use the indices in the following code, because the column labels will be different for each type
  # of model. In other words, if we're doing CESe with Q and plotting sigma, the data.frame will come 
  # back with column names of "sigma" and "CI" on the first and second column, respectively. 
  # plot <-  xyplot(data$CI ~ data$sigma, ...)
  # would break if we, instead, wanted CESe with Q and plotting delta.
  # So, instead, we use the column indices to pick off the correct variables.
  # data[ , 1] contains the the parameter; 
  # data[ , 2] contains the CI width; 
  # data[ , 3] contains the descriptor for the type of graph it is.
  plot <- xyplot(CI ~ param | factor, data=data,
                 scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                             tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                             alternating=FALSE, # eliminates left-right, top-bot alternating of axes
                             x=list(at=xTics),
                             y=list(at=yTics)
                             ), 
                 xlim=xLimits, 
                 ylim=yLimits, 
                 xlab=list(label=xLabels),
                 ylab=list(label=yLabel),
                 # Set text for factor levels
                 strip = strip.custom(factor.levels=stripLabels, 
                                      bg=stripBG, # Sets background transparent to match the graph itself.
                                      par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                      ),
                 as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                 panel = function(x, y){
                   # Picks up the 4th column containing the country abbreviations.
                   ltext(x=x, y=y, labels=data[ , 4], cex=countryAbbrevScaling) 
                 })
  #return(plot)
  return(diagLinesLattice)
}
@

>>>>

########################
# Creates a lattice plot that has the size of confidence intervals on the vertical axis and parameter
# values on the horizontal axis.
# At this point, this function is NOT a general-purpose function, as the number of permutations
# of model, parameter, energy type, and factor is too varied to be able to turn into a general-purpose 
# graph with little time investment at this point.
# This function makes a lattice plot with the following panels (in the arrangement shown)
#
# gammaCDeq   gammaCDex    gammaCDeu
# deltaCESeq  deltaCESex   deltaCESeu
# sigmaCESeq  sigmaCESex   sigmaCESeu
#
# Parameters and models (gamma for CDe and delta and sigma for CESe) are in rows and energyTypes (q, x, and u) 
# are in columns.
#
# textScaling: the amount by which you want to scale the text on the graph.
##
CIvsParamLattice <- function(textScaling=1.0, countryAbbrevScaling=1.0){
  # Set up plotting parameters that are common to both graphs.
  xTics <- c(0.0, 0.5, 1.0) #x axis tic locations
  yTics <- c(0.0, 1.0, 2.0) #y axis tic locations
  xLimits <- c(-0.2, 1.2) #x axis limits
  yLimits <- c(-0.4, 2.4) #y axis limits
  xLabels <- c("$\\gamma$", "$\\delta$", "$\\sigma$") #x axis labels
  yLabel <- "Width of 95\\% Confidence Interval"
  stripLabels <- c("Cobb-Douglas with $q$", "CES with $q$", "CES with $q$",
                   "Cobb-Douglas with $x$", "CES with $x$", "CES with $x$", 
                   "Cobb-Douglas with $u$", "CES with $u$", "CES with $u$")
  stripBG <- "transparent"

@@

  # Diagonal lines for the sigma-CES-Q panel  
  param <- c(0.0, 1.0, 2.0)
  CI <- c(2.0, 0.0, 2.0)
  factor <- c("sigmaCESeQ", "sigmaCESeQ", "sigmaCESeQ")
  diagLineSigmaCESeQDF <- data.frame(param, CI, factor)
  # Diagonal line for the sigma-CES-X panel  
  factor <- c("sigmaCESeX", "sigmaCESeX", "sigmaCESeX")
  diagLineSigmaCESeXDF <- data.frame(param, CI, factor)
  # Diagonal line for the sigma-CES-U panel  
  factor <- c("sigmaCESeU", "sigmaCESeU", "sigmaCESeU")
  diagLineSigmaCESeUDF <- data.frame(param, CI, factor)

@@

  # Make an xyplot that contains the diagonal lines.
  diagLinesLattice <- xyplot(CI ~ param | factor, data=diagLinesDF,
                             type="l", col = "black", lty=2, #controls line parameters
                             scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                             tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                             alternating=FALSE, # eliminates left-right, top-bot alternating of axes
                             x=list(at=xTics),
                             y=list(at=yTics)
                             ), 
                             xlim=xLimits, 
                             ylim=yLimits, 
                             xlab=list(label=xLabels),
                             ylab=list(label=yLabel),
                             strip = strip.custom(factor.levels=stripLabels, # Set text for factor levels
                                                  bg=stripBG,
                                                  par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                                  ),
                             as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                             )

  # Build a data.frame that has all the necessary data from the models.
  gammaCDeqDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="Q")
  deltaCESeqDF <- CIvsParamDF(model="CESe", param="delta", energyType="Q")
  sigmaCESeqDF <- CIvsParamDF(model="CESe", param="sigma", energyType="Q")
  gammaCDexDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="X")
  deltaCESexDF <- CIvsParamDF(model="CESe", param="delta", energyType="X")
  sigmaCESexDF <- CIvsParamDF(model="CESe", param="sigma", energyType="X")
  gammaCDeuDF  <- CIvsParamDF(model="CDe",  param="gamma", energyType="U")
  deltaCESeuDF <- CIvsParamDF(model="CESe", param="delta", energyType="U")
  sigmaCESeuDF <- CIvsParamDF(model="CESe", param="sigma", energyType="U")
  # Change the name of the first columns (the ones that contain the parameter of interest to us) to "param".
  # Doing this allows us to combine the individual data frames into one large data frame.
  colnames(gammaCDeqDF)[1]  <- "param"
  colnames(deltaCESeqDF)[1] <- "param"
  colnames(sigmaCESeqDF)[1] <- "param"
  colnames(gammaCDexDF)[1]  <- "param"
  colnames(deltaCESexDF)[1] <- "param"
  colnames(sigmaCESexDF)[1] <- "param"
  colnames(gammaCDeuDF)[1]  <- "param"
  colnames(deltaCESeuDF)[1] <- "param"
  colnames(sigmaCESeuDF)[1] <- "param"
  # Create one big data.frame with all of the data.
  data <- rbind(gammaCDeqDF, deltaCESeqDF, sigmaCESeqDF, 
                gammaCDexDF, deltaCESexDF, sigmaCESexDF,
                gammaCDeuDF, deltaCESeuDF, sigmaCESeuDF)
  # Now make the graph.
  dataPlot <- xyplot(CI ~ param | factor, data=data,
                     scales=list(cex=scaleTextSize * textScaling, #controls text size on scales.
                                 tck=scaleTickSize, #controls tick mark length. < 0 for inside the graph.
                                 alternating=FALSE, # eliminates left-right, top-bot alternating of axes
                                 x=list(at=xTics),
                                 y=list(at=yTics)
                                 ), 
                     xlim=xLimits, 
                     ylim=yLimits, 
                     xlab=list(label=xLabels),
                     ylab=list(label=yLabel),
                     # Set text for factor levels
                     strip = strip.custom(factor.levels=stripLabels, 
                                          bg=stripBG, # Sets background transparent to match the graph itself.
                                          par.strip.text=list(cex=textScaling) # Scales text in the strip.
                                          ),
                     as.table = TRUE, #indexing of panels starts in upper left and goes across rows.
                     panel = function(x, y){
                       # Picks up the 4th column containing the country abbreviations.
                       ltext(x=x, y=y, labels=data[ , 4], cex=countryAbbrevScaling) 
                     })
  doubleYPlot <- doubleYScale(diagLinesLattice, dataPlot, add.ylab2=FALSE, use.style=FALSE)
  #return(plot)
  #return(diagLinesLattice)
  return(doubleYPlot)
}
@

### 2013-04-29 EconGrowth.Rnw
Went through all of Caleb's text. Lots of questions remain.

@@

\noindent where the elasticity of substitution between capital ($k$) and labor ($l$) is given by $\sigma_1 = \frac{1}{1+\rho_1}$, the elasticity of substitution between ($kl$) and ($e$) is given by $\sigma = \frac{1}{1+\rho}$, $\delta \in [0, 1]$ describes the importance of ($kl$) relative to ($e$), and $\gamma$ is a fitting parameter, expected to be close to unity. Equation \ref{eq:CESe} augments a standard CES production function (Equation \ref{eq:CES}) with energy using a ($kl$)($e$) nesting arrangement, as is typical in the literature. Equation \ref{eq:CES} is a degenerate form of Equation \ref{eq:CESe} where $\delta = 1$.

@@

Total employment and average hours worked per worker per year for the developed nations (United States, United Kingdom, and Japan) for 1980--2011 were obtained from the United States \citet{BLS:2011a}. Total hours worked per year was calculated by multiplying total employment by average hours worked per worker per year.

@@

Exergy ($X$), expressed in TJ/year, was calculated from the thermal energy ($Q$) values discussed in Section \ref{sec:Q_Data}. For fuels, exergy was calculated using a unitless exergy-to-energy allocation factor ($f_{exergy-allocation}$) from \citet{Ayres:2005kn} and shown in Table \ref{tab:Exergy_Allocation}.

@@

$f_{exergy-allocation}$ is the ratio of chemical exergy to the higher heating value for each fuel. ***** Is this right? Are the $X$ values larger than the $Q$ values? The exergy allocation factor for each fuel was assumed to be constant across all economiess. For renewable and nuclear energy sources, the electricity generated is already pure exergy, so $f_{exergy-allocation}$ is unity.

@@

For calculating the exergy of animal and human muscle power, the food-to-energy conversion efficiency and the ratio of food-end-use to food-intake are considered (see equation \ref{eq:animal_muscle_exergy} and equation \ref{eq:human_muscle_exergy}).

\begin{equation} \label{eq:animal_muscle_exergy}
  X_{animals} = E_{animals} \eta_{food \rightarrow work,animals}
\end{equation}

\begin{equation} \label{eq:human_muscle_exergy}
  X_{workers} = E_{humans} \eta_{food \rightarrow work,workers} f_{enduse-to-intake}
\end{equation}

Where $\eta_\mathrm{food-to-work, animals}$ and $\eta_\mathrm{food-to-work, humans}$ are the animal's and human's efficiency of converting food into energy respectively. For humans, an additional factor of the ratio of food-end-use to food-intake is also applied. This ratio looks at how much of a person's energy can actually be used for physical labor, or muscle work, by taking into account food that's used to simply sustain the body's normal function. The food-to-work efficiency values are taken from Ayres and Warr (AYRES AND WARR, THE ECONOMIC GROWTH ENGINE) and the end-use to intake ratio is taken from Wirsenius (WIRSENIUS, 2000). The same ratio could also be applied to animals, but Wirsenius did not contain any information on these values and the magnitude of the animal muscle exergy is quite small compared to the other fuel sources.

Useful work is a concept first developed by (AYRES AND WARR, 2005). Useful work takes exergy analysis one step further and considers where the energy is going and how efficiencies come into play. To calculate useful work, exergy is broken down into several categories by usage type: high temperature heat, medium temperature heat, low temperature heat, mechanical drive, electricity, light, and muscle work. Each of these categories then has an efficiency applied, as opposed to using an overall effiency for the base exergy value. The rationale behind this is that the efficiency of various industries in an economy will be very different, e.g. the efficiency of a coal fired power plant (high temperature heat source) will be very different from that of a residential heating system (low temperature heat source).

Useful work is evaluated in TJ/year, and is given be equation \ref{eq:useful_work}.

\begin{equation} \label{eq:useful_work}
  u = \sum{\phi_\mathrm{i} \epsilon_\mathrm{i} x}
\end{equation}

Where $\epsilon_\mathrm{i}$ is the efficiency of each useful work type and $\phi_\mathrm{i}$ is the allocation factor, where $0 \le \phi_\mathrm{i} \le 1$.

Data on $\epsilon$ and $\phi$ values were taken from the REXS database compiled by Warr (WARR, REXS, 2012). These values are not constant over time, as power generation technologies change and become more efficient. Useful work was only analyzed for the developed economies (the United States, the United Kingdom, and Japan) from 1980--2011, because of the difficulty associated with determining $\phi$ and $\epsilon$ values. In order to determine these parameters, information is needed about the efficiency of different technologies and what industries/areas is consuming the most energy within an economy. This proved to be quite difficult for the developing, petrol, and Malthusian economies so we restricted this part of the analysis to using the data already compiled in the REXS database.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cobb-Douglas Models Without Energy (CD)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

% From http://economics.utoronto.ca/osborne/latex/BIBTEX.HTM
\def\citeapos#1{\citeauthor{#1}'s (\citeyear{#1})} % Posessive citations.

@@

\noindent where the elasticity of substitution between capital ($k$) and labor ($l$) is given by $\sigma_1 = \frac{1}{1+\rho_1}$, the elasticity of substitution between ($kl$) and ($e$) is given by $\sigma = \frac{1}{1+\rho}$, $\delta \in [0, 1]$ describes the importance of ($kl$) relative to ($e$), and $\gamma$ is a fitting parameter, expected to be close to unity. Equation \ref{eq:CESe} augments a standard CES production function (Equation \ref{eq:CES}) with energy using a ($kl$)($e$) nesting structure, as is typical in the literature. Equation \ref{eq:CES} is a degenerate form of Equation \ref{eq:CESe} where $\delta = 1$.

@@

Total employment and average hours worked per worker per year for the developed nations (United States, United Kingdom, and Japan) for 1980--2011 were obtained from the United States \citet{BLS:2013a}. Total hours worked per year was calculated by multiplying total employment by average hours worked per worker per year.

@@

Exergy ($X$), expressed in TJ/year, was calculated from the thermal energy ($Q$) values discussed in Section \ref{sec:Q_Data}. For fuels, exergy was calculated by a unitless exergy-to-energy allocation factor ($f_{exergy-allocation}$) from \citet{Ayres:2005kn} and shown in Table \ref{tab:Exergy_Allocation}. $f_{exergy-allocation}$ is the ratio of chemical exergy to the higher heating value for each fuel. ***** Is this right? Are the $X$ values larger than the $Q$ values? ******* $f_{exergy-allocation}$ for each fuel was assumed to be constant through time across all economies. For renewable and nuclear energy sources, the electricity generated is already pure exergy, so $f_{exergy-allocation}$ is unity.

@@

To calculate the exergy of animal and human muscle power, the food-to-work conversion efficiency ($\eta_{food \rightarrow work}$) and the ratio of food end-use to food intake ($f_{enduse \rightarrow intake}$) were included. $\eta_{food \rightarrow work, animals}$ and $\eta_{food \rightarrow work, workers}$ are animal and worker efficiency of converting food into energy respectively. The values for $\eta_{food \rightarrow work, animals}$ = ***** and $\eta_{food \rightarrow work, workers}$ = ***** were taken from \citet{Ayres:2010ug}. For workers, an additional factor ($f_{enduse \rightarrow intake}$ = ********), which accounts for food energy that is used to sustain the body's normal function as opposed to providing energy for physical labor \citep{Wirsenius:2000vi}. A similar ratio could also be applied to animal work, but it was unavailable from \citet{Wirsenius:2000vi}.\footnote{The error in neglecting $f_{enduse \rightarrow intake}$ for animals is very small, because animal muscle exergy is a small fraction of the total energy supply to any economy, with a maximum of **** \% for ****** in ****** year ******.}

\begin{equation} \label{eq:animal_muscle_exergy}
  X_{animals} = \eta_{food \rightarrow work,animals} E_{animals}
\end{equation}

\begin{equation} \label{eq:human_muscle_exergy}
  X_{workers} = \eta_{food \rightarrow work,workers} f_{enduse \rightarrow intake} E_{workers}
\end{equation}

\noindent

The conversion of a raw energy carrier (e.g., coal) into productive energy for an economy (e.g., electricity) necessarily involves inefficiencies. In the case of coal consumption, only about one-third ****** Is this the number we want to use? ****** of the heat content of coal is converted to useful electricity, the rest is waste heat. The concept of ``useful work,'' pioneered by \citet{Ayres:2005kn}, takes exergy one step further to consider efficiencies of both energy production and end use. To calculate useful work ($U$), each source of exergy (coal, oil and petroleum products, dry natural gas and natural gas plant liquids, biomass/fuel wood, nuclear, solar, wind, and human and animal muscle work) is decomposed into several end uses: high temperature heat (for electricity production **** anything else? ****), medium temperature heat (**** for ???? ****), low temperature heat (for residential heating **** anything else? ****), mechanical drive (**** for ???? ****), electricity (**** for ???? ****), light, and muscle work. The efficiency of each combination of energy source and end use in an economy can be very different, e.g. the efficiency of a coal fired power plant (high temperature heat source, $\sim$30\%) will be very different from a residential heating system (low temperature heat source, $\sim$****\%). Efficiencies specific to each combination of energy source and end use were applied to exergy values to obtain useful work ($U$) input to an economy with the following equation:

\begin{equation} \label{eq:useful_work}
  U = \sum{\phi_i \varepsilon_i X}
\end{equation}

***** Shouldn't this be a double-sum? In other words, shouldn't $X$ have a subscript, too? We sum the U's for all energy sources to get the total useful work for an economy, right? Also, the allocation factor should be specific to the COMBINATION of energy source and end use, right? *****

\noindent where the subscript $i$ indicates the energy end use, $\varepsilon_i$ is the efficiency of each end use, and $\phi_i$ gives the fraction of each energy source that goes to the end use, where $0 \le \phi_i \le 1$. ****** Can we list the raw energy sources and the end uses? *****

Values of $\varepsilon_i$ and $\phi_i$ were calculated from the REXS database supplied by \citet{Warr:2012a}. Values of $X$ were calculated by methods discussed in Section \ref{sec:X_Data}. Most $\varepsilon$ values increase over time **** Are there any that don't? ****, as power generation technologies improve. **** Are there any values of $\varepsilon_i$ that are not increasing with time? **** For most countries, values of $\phi_i$ are not constant with time, as economic sectors with different distributions of raw energy inputs grow at different rates. \citeapos{Warr:2012a} REXS database contains information for developed economies (United States, United Kingdom, and Japan) from 1980--2011 ****** Up to 2011? Isn't this 2000? ******. Extending the REXS database to the BRICS, petrol, and LDC economies is beyond the scope of this work, because longitudinal data for the efficiency of different end uses ($\varepsilon_i$) and the distribution of energy sources to end uses ($\phi_i$) are extremely difficult to find.

**** Do we want to graph the $\varepsilon_i$ and $\phi_i$ values over time?

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Cobb-Douglas Model Without Energy (CD)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
